A Tidy Framework and Infrastructure to Systematically Assemble Spatio-temporal Indexes from Multivariate Data

### 2023-08-23 tidyindex.qmd
typo

| Module                  | Input                                                                   | Operation                           | Output                                                                        |
|---------------|---------|---------------|--------------------------|
| Temporal processing     | $x_{sj}(t)$                                                             | $f[x_{sj}(t)]$                      | $x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}$ |
| Spatial processing      | $x_{tj}(s)$                                                             | $g[x_{tj}(s)]$                      | $x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}$                 |
| Variable transformation | $x_{j}(s; t)$                                                           | $T[x_j(s;t)]$                       | $x^{\text{Trans}}_j(s;t)$                                                     |
| Scaling                 | $x_j(s; t)$                                                             | $[x_j(s;t) - \alpha]/\gamma$        | $x^{\text{Scale}}_j(s;t)$                                                     |
| Dimension reduction     | $\mathbf{x}(s;t)$ | $h[\mathbf{x}(s;t)]$                | $\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < p$      |
| Distribution fit        | $x_j(s; t)$                                                             | $F[x_j(s;t)]$ | $P_j(s;t) \quad P(.) \in [0, 1]$                   |
| Normalising             | $x_j(s; t)$                                                             | $\Phi^{-1}[x_j(s; t)]$              | $z_j(s; t)$                              |
| Benchmarking            | $x_j(s; t)$                                                             | $u[x_j(s;t)]$                       | $b_j(s;t)$                                                                    |
| Simplification          | $x_j(s; t)$                                                             | $v[x_j(s;t)]$                       | $A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}$                                      |

@@

Distribution fit applies the Cumulative Distribution Function (CDF) $F$ of a distribution on the variable $x_j(s; t)$ to obtain the probability values $P_j(s;t) \in [0, 1]$. In SPEI, many distributions, including log-logistic, Pearson III, lognormal, and general extreme distribution, are candidates for the aggregated series. Different fitting methods and different goodness of fit tests may be used to compare the distribution choice on the index value. This could be considered to be a variable transformation because it is usually conducted separately for each variable.  However, very occasionally a fit is conducted on two or more variables simulataneously. For this reaosn, and because it usually is applied later in the pipeline it is neater to make this a separate module.

@@

Benchmark sets a value $b_j(s,t)$ for comparing against the original variable $x_j(s;t)$. This benchmark can be a fixed value consistently across space and time, perhaps extracted from expert knowledge, or determined by the data through the function $u[x_j(s;t)]$. Once a benchmark is set, observations can be highlighted for adjustments in other modules or can serve as targets for monitoring and planning.

>>>>

| Module                  | Input                                                                   | Operation                           | Output                                                                        |
|--------------|---------|---------------|--------------------------|
| Temporal processing     | $x_{sj}(t)$                                                             | $f[x_{sj}(t)]$                      | $x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}$ |
| Spatial processing      | $x_{tj}(s)$                                                             | $g[x_{tj}(s)]$                      | $x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}$                 |
| Variable transformation | $x_{j}(s; t)$                                                           | $T[x_j(s;t)]$                       | $x^{\text{Trans}}_j(s;t)$                                                     |
| Scaling                 | $x_j(s; t)$                                                             | $[x_j(s;t) - \alpha]/\gamma$        | $x^{\text{Scale}}_j(s;t)$                                                     |
| Dimension reduction     | $\mathbf{x}(s;t)$ | $h[\mathbf{x}(s;t)]$                | $\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < p$      |
| Distribution fit        | $x_j(s; t)$                                                             | $F[x_j(s;t)]$ | $P_j(s;t) \quad P(.) \in [0, 1]$                   |
| Normalising             | $x_j(s; t)$                                                             | $\Phi^{-1}[x_j(s; t)]$              | $z_j(s; t)$                              |
| Benchmarking            | $x_j(s; t)$                                                             | $u[x_j(s;t)]$                       | $b_j(s;t)$                                                                    |
| Simplification          | $x_j(s; t)$                                                             | $v[x_j(s;t)]$                       | $A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}$                                      |

@@

Distribution fit applies the Cumulative Distribution Function (CDF) $F$ of a distribution on the variable $x_j(s; t)$ to obtain the probability values $P_j(s;t) \in [0, 1]$. In SPEI, many distributions, including log-logistic, Pearson III, lognormal, and general extreme distribution, are candidates for the aggregated series. Different fitting methods and different goodness of fit tests may be used to compare the distribution choice on the index value. This could be considered to be a variable transformation because it is usually conducted separately for each variable.  However, very occasionally a fit is conducted on two or more variables simultaneously. For this reason, and because it usually is applied later in the pipeline it is neater to make this a separate module.

@@

Benchmark sets a value $b_j(s,t)$ for comparing against the original variable $x_j(s;t)$. This benchmark can be a fixed value consistently across space and time, perhaps extracted from expert knowledge, perhaps extracted from expert knowledge, or determined by the data through the function $u[x_j(s;t)]$. Once a benchmark is set, observations can be highlighted for adjustments in other modules or can serve as targets for monitoring and planning.

### 2023-08-23 tidyindex.tex
typo

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, boxrule=0pt, enhanced, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable]}{\end{tcolorbox}}\fi

@@

\hypertarget{tbl-notation}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2308}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1385}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2308}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.4000}}@{}}
\caption{\label{tbl-notation}Summary of the notation for input,
operation, and output of each pipeline module.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
\(x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}\) \\
Spatial processing & \(x_{tj}(s)\) & \(g[x_{tj}(s)]\) &
\(x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}\) \\
Variable transformation & \(x_{j}(s; t)\) & \(T[x_j(s;t)]\) &
\(x^{\text{Trans}}_j(s;t)\) \\
Scaling & \(x_j(s; t)\) & \([x_j(s;t) - \alpha]/\gamma\) &
\(x^{\text{Scale}}_j(s;t)\) \\
Dimension reduction & \(\mathbf{x}(s;t)\) & \(h[\mathbf{x}(s;t)]\) &
\(\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < p\) \\
Distribution fit & \(x_j(s; t)\) & \(F[x_j(s;t)]\) &
\(P_j(s;t) \quad P(.) \in [0, 1]\) \\
Normalising & \(x_j(s; t)\) & \(\Phi^{-1}[x_j(s; t)]\) &
\(z_j(s; t)\) \\
Benchmarking & \(x_j(s; t)\) & \(u[x_j(s;t)]\) & \(b_j(s;t)\) \\
Simplification & \(x_j(s; t)\) & \(v[x_j(s;t)]\) &
\(A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}\) \\
\end{longtable}

@@

Distribution fit applies the Cumulative Distribution Function (CDF)
\(F\) of a distribution on the variable \(x_j(s; t)\) to obtain the
probability values \(P_j(s;t) \in [0, 1]\). In SPEI, many distributions,
including log-logistic, Pearson III, lognormal, and general extreme
distribution, are candidates for the aggregated series. Different
fitting methods and different goodness of fit tests may be used to
compare the distribution choice on the index value.

@@

Benchmark sets a value \(b_j(s,t)\) for comparing against the original
variable \(x_j(s;t)\). This benchmark can be a fixed value consistently
across space and time or determined by the data through the function
\(u[x_j(s;t)]\). Once a benchmark is set, observations can be
highlighted for adjustments in other modules or can serve as targets for
monitoring and planning.

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@monash.edu}{\nolinkurl{huize.zhang@monash.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Econometrics and Business
Statistics, Monash University, Melbourne,
Victoria, Australia\\ \textsuperscript{2} Institute of
Statistics, University of Natural Resources and Life
Sciences, Vienna, Austria\\ \textsuperscript{3} Department of
Mathematical Sciences, BNU-HKBU United International College, Zhuhai,
Guangdong, China
\begin{abstract}
Indexes are useful for summarizing multivariate information into single
metrics for monitoring, communicating, and decision-making. While most
work has focused on defining new indexes for specific purposes, more
attention needs to be directed towards making it possible to understand
index behavior in different data conditions, and to determine how their
structure affects their values and variation in values. Here we discuss
a modular data pipeline recommendation to assemble indexes. It is
universally applicable to index computation and allows investigation of
index behavior as part of the development procedure. One can compute
indexes with different parameter choices, adjust steps in the index
definition by adding, removing, and swapping them to experiment with
various index designs, calculate uncertainty measures, and assess
indexes' robustness. The paper presents three examples to illustrate the
pipeline framework usage: comparison of two different indexes designed
to monitor the spatio-temporal distribution of drought in Queensland,
Australia; the effect of dimension reduction choices on the Global
Gender Gap Index (GGGI) on countries' ranking; and how to calculate
bootstrap confidence intervals for the Standardized Precipitation Index
(SPI). The methods are supported by a new R package, called
\texttt{tidyindex}.
\end{abstract}
\begin{keywords}
\def\sep{;\ }
indexes\sep data pipeline\sep software design\sep uncertainty\sep 
decision-making
\end{keywords}
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt, enhanced, breakable, interior hidden, frame hidden]}{\end{tcolorbox}}\fi

@@

\hypertarget{tbl-notation}{}
\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2188}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.1406}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.2344}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 6\tabcolsep) * \real{0.4062}}@{}}
\caption{\label{tbl-notation}Summary of the notation for input,
operation, and output of each pipeline module.}\tabularnewline
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endfirsthead
\toprule()
\begin{minipage}[b]{\linewidth}\raggedright
Module
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Input
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Operation
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Output
\end{minipage} \\
\midrule()
\endhead
Temporal processing & \(x_{sj}(t)\) & \(f[x_{sj}(t)]\) &
\(x^{\text{Temp}}_{sj}(t^\prime) \quad t^\prime \in D_{t^\prime}\) \\
Spatial processing & \(x_{tj}(s)\) & \(g[x_{tj}(s)]\) &
\(x^{\text{Spat}}_{tj}(s^\prime) \quad s^\prime \in D_{s^\prime}\) \\
Variable transformation & \(x_{j}(s; t)\) & \(T[x_j(s;t)]\) &
\(x^{\text{Trans}}_j(s;t)\) \\
Scaling & \(x_j(s; t)\) & \([x_j(s;t) - \alpha]/\gamma\) &
\(x^{\text{Scale}}_j(s;t)\) \\
Dimension reduction & \(\mathbf{x}(s;t)\) & \(h[\mathbf{x}(s;t)]\) &
\(\mathbf{y}(s;t) \quad \mathbf{y} \subseteq \mathbb{R}^d, d < p\) \\
Distribution fit & \(x_j(s; t)\) & \(F[x_j(s;t)]\) &
\(P_j(s;t) \quad P(.) \in [0, 1]\) \\
Normalising & \(x_j(s; t)\) & \(\Phi^{-1}[x_j(s; t)]\) &
\(z_j(s; t)\) \\
Benchmarking & \(x_j(s; t)\) & \(u[x_j(s;t)]\) & \(b_j(s;t)\) \\
Simplification & \(x_j(s; t)\) & \(v[x_j(s;t)]\) &
\(A_j(s;t) \in \{a_1, a_2, \cdots, a_z\}\) \\
\bottomrule()
\end{longtable}

@@

Distribution fit applies the Cumulative Distribution Function (CDF)
\(F\) of a distribution on the variable \(x_j(s; t)\) to obtain the
probability values \(P_j(s;t) \in [0, 1]\). In SPEI, many distributions,
including log-logistic, Pearson III, lognormal, and general extreme
distribution, are candidates for the aggregated series. Different
fitting methods and different goodness of fit tests may be used to
compare the distribution choice on the index value. This could be
considered to be a variable transformation because it is usually
conducted separately for each variable. However, very occasionally a fit
is conducted on two or more variables simultaneously. For this reason,
and because it usually is applied later in the pipeline it is neater to
make this a separate module.

@@

Benchmark sets a value \(b_j(s,t)\) for comparing against the original
variable \(x_j(s;t)\). This benchmark can be a fixed value consistently
across space and time, perhaps extracted from expert knowledge, perhaps
extracted from expert knowledge, or determined by the data through the
function \(u[x_j(s;t)]\). Once a benchmark is set, observations can be
highlighted for adjustments in other modules or can serve as targets for
monitoring and planning.
